\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 progressssssslowwwwwwwww

Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}. The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target. 
Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}. In other words, line 3 could be replaced by inserting the following  statement on line 9:
\vspace{-0.2cm}
\begin{verbbox}
map = ace.fn(clx.base, clx.opencl)(map)
\end{verbbox}
\begin{figure}[h!]
\centering
\theverbbox
\end{figure}
\vspace{-0.2cm}

@@

\todo{move this to related work??}Type assignment to generic functions is similar in some ways to template specialization in C++. In effect, both the template header and type parameters at call sites are being generated automatically by Ace. This significantly simplifies a sophisticated feature of C++ and introduces it to targets like OpenCL and C, which do not support templates. Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.

@@

For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising. This simplicity is enabled largely due to  implicit tracking of types throughout the code. First, \verb|numpy| keeps track of the type, shape and order of its arrays. The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,)| and its order is row-major by default. The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size. Our wrapper of \verb|pyopencl| retains this information in the Python representations of the device arrays, \verb|d_input| and \verb|d_output|. An Ace type parameterized can use this metadata to typecheck and generate code for operations like indexing in Ace functions\todo{rephrase}.

@@

The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend. OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner. If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code. Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques. Higher order functions are still not available, and must be simulated by string manipulation. That approach also does not permit the use any of the language extensions that Ace enables (in this case, the only interesting use of the extension mechanism was by the Ace type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec. \ref{examples} for more interesting possibilities).

@@

\section{Extensibility}\label{att}
The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active bases, active types and active targets.  The latter three comprise Ace's extension mechanism. Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core. Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.

\subsection{Active Types}\label{atypes}
Active types are the primary means for extending Ace, and motivate active bases, so we will describe them first. An active type, as mentioned above, is an instance of a class implementing the \verb|ace.ActiveType| interface. Listing \ref{pointers} shows an example of such a class: the \verb|clx.Ptr| class used in Listing \ref{compscript}, which implements the logic of OpenCL's pointer types. The class takes two arguments in its constructor: an address space and the target type of the pointer.
OpenCL pointers support several operations: they can be indexed by an integer type to extract the value being pointed to; one can perform pointer arithmetic on them; and two pointers can be subtracted to produce a value of type \verb|ptrdiff_t| (a device-dependent integer type) if they are in the same address space \cite{opencl11}.

In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation. For example, in a functional language, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, implementing the logic described above in the cases for subscript (\verb|e1[e2]|) and addition (\verb|e1 + e2|). The visitor pattern typically used in object-oriented languages implements essentially the same protocol. This does not work for an extensible language because new cases cannot be added to the types and functions (in a modular, safely composable manner).

Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}. When encountering a compound expression (e.g. \verb|e1[e2]|), Ace defers control over typechecking and translation to the active type of a designated subexpression. The choice of subexpression is determined by Ace's fixed \emph{dispatch protocol}, which is designed taking into consideration Python's semantics for the same form. Below are examples of these choices. Due to space constraints, we do not show the full protocol.
\begin{itemize}
\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g. \verb|-e|) is handed to the type recursively assigned to \texttt{e}.
\item Responsibility over {\bf binary operations} (e.g. \verb|e1 + e2|) is first handed to the type assigned to the left operand. If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call. {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec. \ref{related} for a discussion.}
\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base). If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
\end{itemize}

\subsubsection{Typechecking}
When typechecking a compound expression or statement, the Ace compiler calls a method of the form \verb|type_|$X$ on the active type that has responsibility over it according to the dispatch protocol, where $X$ is typically the name of the syntactic form, taken from the Python grammar \cite{pythonast}. For example, the typing logic for the \verb|Subscript| and \verb|BinOp| forms when operating on OpenCL pointers are shown beginning on lines \ref{pointers}.8 and \ref{pointers}.20, respectively.  The typing methods are provided a \verb|context| and the abstract syntax tree of the \verb|node| being considered. The context is an instance of \verb|ace.Context| and tracks information like variable types, as coordinated by the base. It also includes a method \verb|type| that can be used to assign  type to subexpressions as needed. For example, line \ref{pointers}.8 checks if the type of the slice is an integer type. \verb|Integer| is simply another class in the \verb|clx| library that implements the \verb|ace.ActiveType| interface. Specific integer types like \verb|int| and \verb|ptrdiff_t| are instances of this class (details not shown). Note that the subexpressions are not \emph{automatically} assigned types. This is useful because it allows types to contextually reinterpret them to support more complex constructs (e.g. pattern matching on datatypes, see Sec. \ref{datatypes}).

If a type error is detected, \verb|ace.TypeError| can be raised with an appropriate error message and a reference to the node that produced it (so that Ace can extract line and column numbers). We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}. A common frustration with general-purpose abstraction mechanisms is that they can produce  verbose and cryptic error messages revealing implementation details of the abstraction being encoded (e.g. template metaprogramming \cite{template-metaprogramming}).

\begin{codelisting}
\lstinputlisting{listing9.py}
\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
\label{pointers}
\end{codelisting}
\subsubsection{Translation}
Once typechecking a method is complete, the compiler must subsequently translate each Ace source expression into an expression in the target language. This has been OpenCL in the examples thus far, but we describe a generalization of this in the next section.

@@

@@

If no active target is explicitly provided, as in our example, the active base  provides a default target. The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries. One base can support multiple targets. We will discuss targets in Sec. \ref{targets}.

Thus far, we have been discussing the \verb|OpenCL| module in our examples. This module faithfully implements all the types and operations of the OpenCL kernel language, a portable standard for writing low-level code on multi-core processors and accelerators \cite{opencl11}. The functions we wrote have used these primitives with the \verb|OpenCL.OpenCL| backend, so the translation has been direct and no run-time overhead has been introduced. Thus, as described so far, Ace has affirmatively resolved the question it was originally conceived to address, discussed in Section 2.1.

>>>>

Generic functions are represented at compile-time as instances of \verb|ace.GenericFn| and consist of an abstract syntax tree, an {active base} and an {active target}. The purpose of the \emph{decorator} on line 3 is to replace the Python function on lines 4-8 with an Ace generic function having the same syntax tree and the provided active base and active target. 
Decorators in Python are simply syntactic sugar for applying the decorator function directly to the function  being decorated \cite{python}. In other words, line 3 could be replaced by inserting the following  statement on line 9:
\vspace{-0.28cm}
\begin{verbbox}
map = ace.fn(clx.base, clx.opencl)(map)
\end{verbbox}
\begin{figure}[h!]
\centering
\theverbbox
\end{figure}
\vspace{-0.28cm}

@@

Type assignment to generic functions is similar in some ways to template specialization in C++. In effect, both a template header and type parameters at call sites are being generated automatically by Ace. This simplifies a sophisticated feature of C++ and enables its use with other targets like OpenCL. %Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.

@@

For developers experienced with the usual OpenCL or CUDA workflow, the fact that this can be accomplished in a total of 6 statements may be surprising. This simplicity is possibly largely due to  implicit tracking of types throughout the code. First, \verb|numpy| keeps track of the type, shape and order of its arrays. The type of \verb|input|, for example, is \verb|numpy.double| by default, its shape is \verb|(1024,1024)| and its order is row-major by default. The \verb|pyopencl| library that our mechanism is built atop uses this information to automatically call the underlying OpenCL host API function for transferring byte arrays to the device without requiring the user to calculate the size. Our wrapper of \verb|pyopencl| further retains this information in the Python wrappers around the device arrays, \verb|d_input| and \verb|d_output|. The Ace active type of such wrappers can be designated to be an instance of \verb|clx.NPArray| parametermized by this metadata. This type knows how to typecheck and translate operations like indexing automatically (see Sec. \ref{atypes}).

@@

The same program written using the OpenCL C API directly is an order of magnitude longer and significantly more difficult to comprehend. OpenCL not support higher-order functions nor is there any way to write \verb|map| in a type-generic manner. If we instead use the \verb|pyopencl| library and apply the techniques described in \cite{klockner2011pycuda}, the program is still twice as large and less readable than this code. Both the \verb|map| and \verb|negate| functions must be explicitly specialized with the appropriate types using string manipulation techniques. Higher order functions are still not available, and must also be simulated by string manipulation. That approach also does not permit the use any of the language extensions that Ace enables (e.g. the type for \verb|numpy| arrays, which ensures that indexing respects ordering; see Sec. \ref{examples} for more interesting possibilities).

@@

\section{Extensibility}\label{att}
The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.  The latter three comprise Ace's extension mechanism. Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core. %Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.

\subsection{Active Types}\label{atypes}
Active types are the primary means for extending Ace with new abstractions. An active type, as mentioned previously, is an instance of a class implementing the \verb|ace.ActiveType| interface. Listing \ref{cplx} shows an example of such a class: the \verb|clx.Cplx| class used in Listing \ref{compscript}, which implements the logic of complex numbers. The constructor takes as a parameter any numeric type in \verb|clx| (line \ref{cplx}.2).

\subsubsection{Dispatch Protocol}
\begin{codelisting}
\lstinputlisting{listing9.py}
\caption{\texttt{[}in \texttt{examples/clx.py]} The active type family \texttt{Ptr} implements the semantics of OpenCL pointer types.}
\label{cplx}
\end{codelisting}
In a compiler for a monolithic language, there would be a \emph{syntax-directed} protocol governing typechecking and translation. In a compiler written in a functional language, for example, one would declare datatypes that captured all forms of types and expressions, and the typechecker would perform exhaustive case analysis over the expression forms, so all the semantics would be implemented in one place. The visitor pattern typically used in object-oriented languages implements essentially the same protocol. This does not work for an extensible language because new cases and logic need to be added modularly, in a safely composable manner.

Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}. When encountering a compound term (e.g. \verb|e1[e2]|), the compiler defers control over typechecking and translation to the active type of a designated subexpression (e.g. \verb|e1|) determined by Ace's fixed \emph{dispatch protocol}. Below are examples of these choices. Due to space constraints, we do not show the full protocol.
\begin{itemize}
\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g. \verb|-e|) is handed to the type recursively assigned to \texttt{e}.
\item Responsibility over {\bf binary operations} (e.g. \verb|e1 + e2|) is first handed to the type assigned to the left operand. If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call. {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec. \ref{related} for a discussion.}
\item Responsibility over \textbf{constructor calls} (\verb|[t](e1, ..., en)|), where \verb|t| is a \emph{compile-time Python expression} evaluating to an active type, is handed to that type. If \verb|t| evaluates to a \emph{family} of types, like \verb|clx.Cplx|, the active type is first generated via a class method, as discussed below.
%\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base). If this type does not provide special assignment semantics, the base must handle it.%Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
\end{itemize}

\subsubsection{Typechecking}
When typechecking a compound expression or statement, the Ace compiler temporarily hands control to the object selected by the dispatch protocol by calling the method \verb|type_|$X$, where $X$ is  the name of the syntactic form, taken from the Python grammar \cite{pythonast} (appended with a suffix in some cases).

For example, if \verb|c| is a complex number, then \verb|c.ni| and \verb|c.i| are its non-imaginary and imaginary components, respectively. These expressions are of the form \verb|Attribute|, so the typechecker calls \verb|type_Attribute| (line \ref{cplx}.7).
This method receives the compilation context, \verb|context|, and the abstract syntax tree of the expression, \verb|node| and must return a type assignment for the node, or raise an \verb|ace.TypeError| if there is an error. In this case, a type assignment is possible if the attribute name is either \verb|"ni"| or \verb|"i"|, and an error is raised otherwise (lines \ref{cplx}.8-\ref{cplx}.10). We note that error messages are an important and sometimes overlooked facet of {ease-of-use} \cite{marceau2011measuring}. A common frustration with using general-purpose abstraction mechanisms to encode an abstraction is that they can produce  verbose and cryptic error messages that reflect the implementation details, rather than the semantics. %Active types permit direct encoding of semantics and customization of error messages.

Complex numbers also support binary arithmetic operations partnered with both other complex numbers and with non-complex numbers, treating them as if their imaginary component is zero. The typechecking rules for this logic is implemented on lines \ref{cplx}.17-\ref{cplx}.29. Because arithmetic operations are meant to be symmetric, the dispatch protocol checks both subexpressions for support, favoring the left to ensure that the semantics remain deterministic. In either position, the implementation begins by recursively assigning a type to the other operand in the current context via the \verb|context.type| method (line \ref{cplx}.24). If supported, it applies the C99 rules for arithmetic operations to determine the resulting type (\cite{c99}, not shown).

Finally, a complex number can be constructed inside an Ace function using Ace's special constructor form: \verb|[clx.Cplx](3,4)| represents $3+4i$, for example. The term within the braces is evaluated at \emph{compile-time}. Because \verb|clx.Cplx| evaluates not to an active type, but to a class, this form is assigned a type by handing control to the class object via the \emph{class method} \verb|type_New|. It operates as expected, extracting the types of the two arguments to construct an appropriate complex number type (lines \ref{cplx}.50-\ref{cplx}.57), raising a type error if the arguments cannot be promoted to a common type according to the rules of C99 or if two arguments have not been provided (an exercise for the reader: modify this method to also allow a single argument for when the imaginary part is 0).

%Note that in each example, subexpressions are not \emph{automatically} assigned types. Types must be requested from the context. This is useful because it allows types to reinterpret subexpressions of terms they control, to supporting more complex constructs (e.g. pattern matching on functional datatypes, see Sec. \ref{datatypes}).

\subsubsection{Translation}
Once typechecking a method is complete, the compiler enters the translation phase, where terms in the target language are generated from Ace terms. Terms in the target language are generated by calling methods of the \emph{active targets}. The translation phase operates similarly to typechecking, using the dispatch protocol to invoke methods named \verb|trans_|$X$. These methods have access to the context and node just as during typechecking, as well as the target.

Translation of terms of the form \verb|c.ni| and \verb|c.i| operates via \verb|trans_Attribute|. Because we are implementing complex numbers as vector types in OpenCL, we generate an access to the left element, accessed via the attribute \verb|s0|, to access the non-imaginary component and and \verb|s1| to access the imaginary component. The target provides a programmatic API for generating code in the target language via its methods, as can be seen. Translation can assume typechecking succeeded, so the implementation assumes that if the attribute is not \verb|ni|, it is \verb|i|.

@@

\subsubsection{Target Types}
Every active type must translate to a unique  \emph{target type} -- a type in the target language (dynamically typed target languages can be thought of as only having a single type, \verb|dyn| \cite{pfpl}). Target types are constructed via methods of a provided {active target}, called \verb|target| throughout this listing. Here, a complex number translates to a pair of values, each having the type that is the target type of \verb|t|, the numeric type that was passed into the constructor (lines \ref{cplx}.7-\ref{cplx}.8). The \verb|tuple| method returns a representation of an actual type (not an active type) in the target language as the return value of the \verb|tuple| method (e.g. 2-tuples of integers compile to the vector type \verb|int2| in OpenCL). This design still allows the complex number type being defined here to supports any target for which tuples can be defined directly however (e.g. CUDA, Python, etc.).

\todo{nathan's work on representational consistency?}

@@
